---
title: "Cleaning data for Data challenge"
format:
  html:
    theme: flatly
    code-fold: true
---

```{r echo = F, warning = F, message = F}
library(tidyverse)
library(ggplot2)
library(readxl)
library(readr)
```

:::{.callout-tip  collapse="false"} 
Epi calendar from 31/12/2021 to 31/12/2023

```{r warning = F, message = F}
# Calendar used for hospitalised data
calendar <- read_csv("https://raw.githubusercontent.com/IgnaLeiva/sanofii/main/doc/calendar2022.csv")
calendar$date <- as.Date(calendar$date, "%d/%m/%Y")

# Calendar upload later

calendar2 <- read_csv("/Users/igna/Documents/LSHTM/Term 2/Data Challenge/lshtm-dc-sanofi/data/epi_weeks.csv")

# calendar3 <- data.frame(year = calendar2$year, week = calendar2$epi_wk_no, start = calendar2$epi_dates, end = c(calendar2$epi_dates[-1],1))
# calendar3$end <- calendar3$end -1
```

:::

:::{.callout-tip  collapse="false"}
Population based on INE [Instituto de estadistica](https://www.ine.gob.cl/estadisticas/sociales/demografia-y-vitales/proyecciones-de-poblacion)
::: 

```{r}
population <- read_csv("/data/raw_data/Countries_population.csv")
population <- population %>% 
     filter(code == "CL") %>% 
     select(year, population)
```


## Data set all for COVID-19 hospitalization ages

The data set was taken from the same previous source. It is called "HospitalizadosEtario_Acumulado_std.csv". The number represents the cumulative number of hospitalised people.

There were two measurements per week. I took the last measurement as the representative one. Then to get the "new number (considering admissions and discharges)," I subtracted the prior week's value

```{r warning = F, message = F}
hospilalised <- read_csv("/data/raw_data/CL/HospitalizadosEtario_Acumulado_std.csv")
colnames(hospilalised) <- c("age_group", "sex", "date", "number")

hospilalised$date <- as.Date(hospilalised$date, "%d-%m-%Y")

hospilalised <- (merge(hospilalised, calendar, by = 'date'))
hospilalised$age_group <- as.factor(hospilalised$age_group)
hospilalised$year <- format(hospilalised$date, "%Y")

hospilalised_week <- hospilalised %>% 
                      group_by(date, year, week, age_group) %>% 
                      summarise("cumul_cases" = sum(number, na.rm = T)) %>% 
                      arrange(date, year, week,age_group) 

hospilalised_week <-  hospilalised_week %>% 
                      group_by(year, week, age_group) %>% 
                      summarise("cumul_cases" = last(cumul_cases)) %>% 
                      filter(cumul_cases > 0) %>% 
                      arrange(year, week,age_group) 
                      
# Change levels 

hospilalised_week$age_group <-as.character(hospilalised_week$age_group)

hospilalised_week$age_group <- ifelse(hospilalised_week$age_group == "00  5 años", "0-5",
       ifelse(hospilalised_week$age_group == "5  17 años", "5-17",
              ifelse(hospilalised_week$age_group == "18  49 años", "18-49",
                     ifelse(hospilalised_week$age_group == "50  59 años", "50-59",
                            ifelse(hospilalised_week$age_group == "60  69 años", "60-69",
                                   ifelse(hospilalised_week$age_group == "70  79 años", "70-79",
                                          ifelse(hospilalised_week$age_group == "80 y más años", "80+",0)))))))

hospilalised_week$age_group <- factor(hospilalised_week$age_group, levels = c("0-5","5-17","18-49","50-59","60-69","70-79", "80+"))
hospilalised_week <- arrange(hospilalised_week,year, week,age_group)

# Cases prior week
hospilalised_week$prior_week <- c(rep(NA,7),hospilalised_week$cumul_cases[-c(379:385)])
hospilalised_week$daily <- hospilalised_week$cumul_cases - hospilalised_week$prior_week

hospilalised_week$denominator <- as.numeric(ifelse(hospilalised_week$year == 2022, population[7,2],
                                        ifelse(hospilalised_week$year == 2023, population[8,2], NA)))
```

## Data from SARI cases (sentinel data)

Data extracted from Epi Department at [Ministry of Health]((http://epi.minsal.cl/visualizacion-interactiva-influenza/#/))

```{r warning = F, message = F}
sari_chile <- read_csv('/data/raw_data/CL/SARI_minsal.csv')
sari_chile <- sari_chile[,c(1:5)]

sari_chile$virus2 <- ifelse(sari_chile$virus %in% c("Influenza A","Influenza B"),"flu",
                            ifelse(sari_chile$virus == "VRS", "RSV", sari_chile$virus ))

sari_chile <- filter(sari_chile, !is.na(virus2))
colnames(sari_chile) <- c("year", "week", "subtipe", "subtipe2", "positive_sample", "virus")


sari_chile <- sari_chile %>% 
                group_by(year, week,virus) %>% 
                summarise("positive_sample" = sum(positive_sample)) 
sari_chile <- pivot_wider(sari_chile, names_from = "virus", values_from = "positive_sample")

# set prior 2020 COVID to cero

sari_chile$COVID[sari_chile$year != 2022 ] <- 0


  
```



### Attaching  population to sari cases 

```{r}
sari_chile$denominator <- as.numeric(ifelse(sari_chile$year == 2016, population[1,2],
                                 ifelse(sari_chile$year == 2017, population[2,2],
                                        ifelse(sari_chile$year == 2018, population[3,2],
                                               ifelse(sari_chile$year == 2019, population[4,2],
                                                      ifelse(sari_chile$year == 2022, population[7,2],0))))))

```

### The data set for all age - FLU - RSV - COVID from Minsal

It was assumed that sentinel data represent 3% the population. So, the denominator was multiplied by 0.03
```{r}
chi_premerged_total <- data.frame(
  id =1:260,
  data_source = "EPI MINSAL",
  country = "CL",
  hemisphere = "SH",
  week = sari_chile$week,
  year = sari_chile$year,
  age_group = "ALL",
  hsp_rate = round(((sari_chile$flu + sari_chile$RSV + sari_chile$COVID)/(sari_chile$denominator*0.03))*100000,2),
  hsp_rate_flu = round((sari_chile$flu/(sari_chile$denominator*0.03))*100000,2),
  hsp_rate_rsv = round((sari_chile$RSV/(sari_chile$denominator*0.03))*100000,2),
  hsp_rate_covid19 = round((sari_chile$COVID/(sari_chile$denominator*0.03))*100000,2),
  cases_rate_flu = NA,
  cases_rate_rsv = NA,
  cases_rate_covid19 = NA,
  hsp_abs = (sari_chile$flu + sari_chile$RSV + sari_chile$COVID),
  hsp_abs_flu = sari_chile$flu,
  hsp_abs_rsv = sari_chile$RSV,
  hsp_abs_covid = sari_chile$COVID,
  cases_abs_flu = NA,
  cases_abs_rsv = NA,
  cases_abs_covid19 = NA,
  denominator = sari_chile$denominator,
  subtype_a_abs = NA,
  subtype_b_abs = NA,
  subtype_c_abs = NA,
  subtype_a_rate = NA,
  subtype_b_rate = NA,
  subtype_c_rate = NA
)


#write_csv(chi_premerged_total, "/Users/igna/Documents/LSHTM/Term 2/Data Challenge/lshtm-dc-sanofi/data/premerged_data/CL_premerged_total.csv")

```

## The data set by age for hospitalizations (only COVID) - From Ministry of Science Git-Hub

```{r}
chi_premerged_by_age <- data.frame(
  id =1:385,
  data_source = "EPI MINSAL - Ministry of Science",
  country = "chi",
  hemisphere = "sh",
  week = hospilalised_week$week,
  year = hospilalised_week$year,
  age_group = hospilalised_week$age_group,
  hsp_rate = NA,
  hsp_rate_flu = NA,
  hsp_rate_rsv = NA,
  hsp_rate_covid19 = round((hospilalised_week$daily/hospilalised_week$denominator)*100000,2),
  cases_rate_flu = NA,
  cases_rate_rsv = NA,
  cases_rate_covid19 = NA,
  hsp_abs = NA,
  hsp_abs_flu = NA,
  hsp_abs_rsv = NA,
  hsp_abs_covid = hospilalised_week$daily,
  cases_abs_flu = NA,
  cases_abs_rsv = NA,
  cases_abs_covid19 = NA,
  denominator = hospilalised_week$denominator,
  subtype_a_abs = NA,
  subtype_b_abs = NA,
  subtype_c_abs = NA,
  subtype_a_rate = NA,
  subtype_b_rate = NA,
  subtype_c_rate = NA
)

chi_premerged_by_age <- filter(chi_premerged_by_age, !is.na(denominator)) # delete 7 first rows

#write_csv(chi_premerged_by_age, "data/premerged_data/chi_premerged_by_age.csv")

```


## Australia

This code was meant to fix and the population error (just the last year)
```{r}

flunet_raw <- read_csv("data/processed_data/AUS/australia_fluNet.csv")


flunet_raw$denominator[flunet_raw$denominator == 25890.8] <- 25978935


aus_premerged_total <- data.frame(
  data_source = flunet_raw$data_source,
  country = "AUS",
  hemisphere = "SH",
  week = flunet_raw$week,
  year = flunet_raw$year,
  age_group = "ALL",
  hsp_rate = NA,
  hsp_rate_flu = round((flunet_raw$hsp_abs_flu/flunet_raw$denominator)*100000,2),
  hsp_rate_rsv = round((flunet_raw$hsp_abs_rsv/flunet_raw$denominator)*100000,2),
  hsp_rate_covid19 = NA,
  cases_rate_flu = NA,
  cases_rate_rsv = NA,
  cases_rate_covid19 = NA,
  hsp_abs = NA,
  hsp_abs_flu = flunet_raw$hsp_abs_flu,
  hsp_abs_rsv = flunet_raw$hsp_abs_rsv,
  hsp_abs_covid = NA,
  cases_abs_flu = NA,
  cases_abs_rsv = NA,
  cases_abs_covid = NA,
  denominator = flunet_raw$denominator,
  subtype_a_abs = flunet_raw$subtype_a_abs,
  subtype_b_abs = flunet_raw$subtype_b_abs,
  subtype_c_abs = NA,
  subtype_a_rate = round((flunet_raw$subtype_a_abs/flunet_raw$denominator)*100000,2),
  subtype_b_rate = round((flunet_raw$subtype_b_abs/flunet_raw$denominator)*100000,2),
  subtype_c_rate = NA
)

#write_csv(aus_premerged_total, "data/premerged_data/AU_premerged_total.csv")

```

## France non_sentinel

This code takes FR data from processed_data folder (those don't have all the needed columns). This code provides the missing columns

```{r}
preclean_fran_flu<- read_csv("data/processed_data/FRA/france_fluNet.csv")

preclean_fran_covid<- read_csv("data/processed_data/FRA/france_covid19.csv")


FR_premerged_total <- data.frame(
  data_source = ifelse(preclean_fran_flu$origin_source == "NONSENTINEL", "FluNet - Non Sentinel","FluNet - Sentinel" ),
  country = "FR",
  hemisphere = "NH",
  week = preclean_fran_flu$week,
  year = preclean_fran_flu$year,
  age_group = "ALL",
  hsp_rate = NA,
  hsp_rate_flu = round((preclean_fran_flu$hsp_abs_flu/preclean_fran_flu$denominator)*100000,2),
  hsp_rate_rsv = round((preclean_fran_flu$hsp_abs_rsv/preclean_fran_flu$denominator)*100000,2),
  hsp_rate_covid19 = NA,
  cases_rate_flu = NA,
  cases_rate_rsv = NA,
  cases_rate_covid19 = NA,
  hsp_abs = NA,
  hsp_abs_flu = preclean_fran_flu$hsp_abs_flu,
  hsp_abs_rsv = preclean_fran_flu$hsp_abs_rsv,
  hsp_abs_covid = NA,
  cases_abs_flu = NA,
  cases_abs_rsv = NA,
  cases_abs_covid = NA,
  denominator = preclean_fran_flu$denominator,
  subtype_a_abs = preclean_fran_flu$subtype_a_abs,
  subtype_b_abs = preclean_fran_flu$subtype_b_abs,
  subtype_c_abs = NA,
  subtype_a_rate = round((preclean_fran_flu$subtype_a_abs/preclean_fran_flu$denominator)*100000,2),
  subtype_b_rate = round((preclean_fran_flu$subtype_b_abs/preclean_fran_flu$denominator)*100000,2),
  subtype_c_rate = NA
)

FR_premerged_total$hsp_abs_covid <- left_join(preclean_fran_flu, preclean_fran_covid, by = c("year", "week"))$hsp_abs_covid19

FR_premerged_total$hsp_rate_covid19 <- round((FR_premerged_total$hsp_abs_covid/FR_premerged_total$denominator)*100000,2)

#write_csv(FR_premerged_total, "data/premerged_data/FR/FR_premerged_total.csv")

```


